Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements

### 2022-02-03 case-studies.Rmd
Revise text on MOMSPI regression analysis
Update main text and SI figure to refer to the gamma-Poisson regression
analysis, plus minor fixes.

Since these analysis are both based on proportions, bias could in principle impact their results.
We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)].
To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \@ref(fig:leopold2020host-variation)).
<!-- Explanation -->
Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't significantly affect the mean efficiency.
<!-- Conclusion -->
Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.

@@

The mean efficiency varies substantially across vaginal samples (Figure \@ref(fig:momspi-mean-efficiency-dist).
This variation appears to be primarily driven by variation in which species is most abundant, samples in which a _Lactobacillus_ species is most abundant (after calibration) typically have an efficiency that is 3-20X greater than samples in which _G. vaginalis_ is most abundant.
Shifts between _Lactobacillus_-dominance and _Gardnerella_-dominance are common in between-women comparisons and occasionally occur between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
These shifts typically result in substantial fold changes in mean efficiency (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).

@@

We next considered whether the observed variation in mean efficiency could cause systematic error in a DA analysis.
In particular, we hypothesized that DA analysis of species proportions versus a covariate that is associated with _Lactobacillus_ and/or _Gardnerella_ would be particularly prone to spurious results.
Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we _a priori_ expected to be associated with the proportions of _Lactobacillus_ and _Gardnerella_.
Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult). @cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.
In addition, it is commonly observed that samples from women with and without BV that are dominated by _Lactobacillus spp._ tend to have higher diversity, whereas samples dominated by _Gardnerella_ tend to have lower diversity.
We therefore chose to perform a DA analysis of species proportion versus alpha diversity, hypothesizing that _Lactobacillus_ and _Gardnerella_ would drive a negative association of mean efficiency with diversity and thereby distort DA estimates for all species.
We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
We then estimated the LFC in proportion from low- to high-diversity samples for a 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles.
As expected, the mean efficiency was higher in low-diversity samples due to a larger fraction of samples dominated by _Lactobacillus_ and a lower fraction with samples dominated by _Gardnerella_.
The decline in mean efficiency in the high diversity group caused a concomitant increase in the LFCs of all species when bias was not accounted for.
This error due to bias resulted in serious magnitude errors in nearly all species and sign errors in over one third, with errors of both types seen for several clinically relevant species.

### Notes

- It might be cleaner story-wise to redo the diversity partitioning using the threshold identified by @cartwright2018mult
- For the DA analysis of proportions vs. diversity, I tried other types of DA analysis methods (Gamma-Poisson regression and regression on ranks, similar to Wilcoxon test) to understand how the impact of bias differs across methods. In each case bias has a big effect but the species whose estimates are most affected varies. I would need to revise and verify this analysis before including its results. One reason for doing so would be to show that our qualitative finding about how bias distorts results applies beyond the basic linear regression model.
- Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.

@@

The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of section X help suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.

>>>>

Since these analysis are both based on proportions, bias could in principle impact their results.
We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for.^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)]
To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \@ref(fig:leopold2020host-variation)).
<!-- Explanation -->
Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't significantly affect the mean efficiency.
<!-- Conclusion -->
Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.

@@

The mean efficiency varies substantially across vaginal samples (Figure \@ref(fig:momspi-mean-efficiency-dist).
This variation appears to be primarily driven by variation in which species is most abundant, samples in which a _Lactobacillus_ species is most abundant (after calibration) typically have an efficiency that is 3-20X greater than samples in which _G. vaginalis_ is most abundant.
Shifts between _Lactobacillus_-dominance and _Gardnerella_-dominance are common in between-women comparisons and occasionally occur between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)).
These shifts typically result in substantial fold changes in mean efficiency (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)) and can cause spurious fold changes in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).
Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance would be more extreme for vaginal microbiome primer sets that fail to amplify _Gardnerella_ (@graspeuntner2018sele).

@@

We next considered whether the observed variation in mean efficiency could cause systematic error in a DA analysis.
In particular, we hypothesized that DA analysis of species proportions versus a covariate that is associated with _Lactobacillus_ and/or _Gardnerella_ would be particularly prone to spurious results.
Patient metadata was not available due to privacy restrictions; we therefore sought a clinically relevant covariate to use in a regression analysis that could be determined directly from the microbiome profiles and that we _a priori_ expected to be associated with the proportions of _Lactobacillus_ and _Gardnerella_.
Alpha diversity metrics such as species richness, the Shannon index, and the (Inverse) Simpson index have been repeatedly found to be strongly positively associated with bacterial vaginosis (BV; @srinivasan2012bact, @cartwright2018mult).^[@cartwright2018mult found that an observed Simpson Diversity Index of 0.82 (corresponding to an order-2 effective number of species of 5.6) classified high diversity samples as BV positive with a sensitivity of 100% and specificity of 85.1%.]
In addition, it is commonly observed that samples from women with and without BV that are dominated by _Lactobacillus spp._ tend to have higher diversity, whereas samples dominated by _Gardnerella_ tend to have lower diversity.
We therefore chose to perform a DA analysis of species proportion versus alpha diversity, hypothesizing that _Lactobacillus_ and _Gardnerella_ would drive a negative association of mean efficiency with diversity and thereby distort DA estimates for all species.
We split samples into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles.
We then estimated the LFC in proportion from low- to high-diversity samples for 30 species passing a basic prevalence threshold, with and without bias correction.
A large fraction of entries in the species count matrix were small counts or zeros.
We used therefore used gamma-Poisson (or negative-binomial) regression of the observed read counts to fit a linear model to log species proportions while accounting for uncertainty from the random sampling of reads.
We accounted for bias by including an offset term in the linear model equal to the log ratio of the species' efficiency to the sample mean efficiency.

The mean efficiency was on average lower in high-diversity samples due to a reduction in _Lactobacillus_ dominance (LFC of -1.6, or a 4.8X reduction). 
Our results from the previous section therefore suggest that accounting for bias should reduce the estimated LFC for each species by about this same amount.
We found that, as expected, bias correction reduced the LFC of each species, though the effect was variable and typically somewhat smaller.
Bias correction led to substantial magnitude changes for several species and sign changes in 3 species; however, the relative difference between the corrected and uncorrected LFCs was modest for the 19/30 species with the largest LFCs of over 2 (corresponding to a FC of 7.4X) even after correction.
These large LFCs occur in species that rarely if ever reach high proportions and so are likely driven by the sum-to-one constraint: The high diversity communities are those not dominated by _Lactobacillus_ spp. (or to a lesser extent _G. vaginalis_ and BVAB1).
Therefore it is not clear how well it will generalize to regressions on other covariates.

@@

The impact of bias can depend on protocol, biological system, and type of DA analysis being done.
Though these case studies span a highly limited range of possibilities, when combined with the theoretical results of Section \@ref(differential-abundance) suggest some general conclusions about how and when bias will impact DA analyses based on fold changes in proportions.

### 2022-02-03 supplemental-figures.Rmd
Revise text on MOMSPI regression analysis
Update main text and SI figure to refer to the gamma-Poisson regression
analysis, plus minor fixes.

```{r momspi-regression, fig.cap = '(ref:cap-momspi-regression)'}
fs::path(
  "notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files", 
  "figure-html5/regression-lm-alt-1.svg"
) %>%
  include_svg
```

(ref:cap-momspi-regression) **Bias distorts log fold changes (LFCs) in species proportions in a regression analysis of vaginal microbiome samples from the MOMS-PI study.** Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles. The LFC in proportion from low- to high-diversity samples was estimated for 30 common species by simple linear regression, using calibrated (bias-corrected) and observed (uncorrected) microbiome profiles following a simple zero-replacement procedure. Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95% confidence intervals for each species. The difference between the calibrated and observed estimate for each species equals the negative LFC in mean efficiency.

>>>>

```{r momspi-regression, fig.cap = '(ref:cap-momspi-regression)'}
fs::path(
  "notebook/_posts/2021-12-15-momspi-regression-diversity/momspi-regression-diversity_files", 
  "figure-html5/regression-gp-1.svg"
) %>%
  include_svg
```

(ref:cap-momspi-regression) **Bias inflates estimated log fold changes (LFCs) in species proportions with diversity in vaginal microbiome samples from the MOMS-PI study.** Samples were split into low, medium, and high diversity groups based on Shannon diversity in observed (uncalibrated) microbiome profiles. The LFC in proportion from low- to high-diversity samples was estimated for 30 common species using gamma-Poisson regression with and without bias correction. Panel A shows the distribution of point estimates; Panel B shows the point estimates and 95% Bayesian credible intervals for each species.
